Optimal Algorithms for Dynamic Balanced Graph Partitioning

### 2020-05-13 paper/disc20clustering.tex
rewriting the LB (fixed)

 \usepackage{balance}
 \usepackage{amsthm,amsmath,array,colortbl,graphicx,multirow}
 \usepackage{comment}
 \usepackage{balance}
 \usepackage{tikz}
 \usepackage{amsmath}
 \usetikzlibrary{patterns} %
 \usepackage{algorithm}
 \usepackage[font={footnotesize}]{subcaption}
 \usepackage[font={footnotesize}]{caption}https://www.wykop.pl/
 \usepackage{breakcites}
 \usepackage{booktabs}
 \usepackage{diagbox}
 \usepackage{xcolor}
 \usepackage{colortbl}
 \usepackage{cleveref}
 \usepackage{enumitem}

@@

\begin{abstract}
	Distributed   applications,  including  batch  processing, streaming, scale-out databases,
	or machine learning, generate a~significant amount of network traffic.
	By collocating frequently communicating nodes (e.g., virtual machines) on the same clusters (e.g., server or rack), we can reduce the network load and  improve application performance. 
	However, the communication pattern of different applications is often unknown a~priori and may change over time, hence it needs to be learned in an online manner.
	%
	This paper revisits the online 
	balanced partitioning problem 
	(introduced by Avin et al.~at DISC 2016)
	that asks for an algorithm that strikes
	an optimal tradeoff between the benefits
	of collocation (i.e., lower network load) 
	and its costs (i.e., migrations). 
	%
	Our first contribution is a~significantly improved
	\maciek{How to put here the information that it is for deterministic algorithms only? This sentence is already very long.}lower bound of $\Omega(k\cdot \ell)$ on the
	competitive ratio, where $\ell$ is the number
	of clusters and $k$ is the cluster size,
	even for a~scenario in which the communication
	pattern is static and can be perfectly partitioned;
	we also provide an asymptotically tight upper bound 
	of $O(k\cdot \ell)$ for this scenario.
	For $k=3$, we contribute an asymptotically tight upper bound
	of $\Theta(\ell)$
	for the general model in which the
	communication pattern can change arbitrarily over time.
	We improve the result for $k=2$ by providing a~strictly $6$-competitive upper bound for the general model.
	In contrast to most prior work, our algorithms respect all capacity constraints and do not require resource augmentation.

@@

\begin{theorem}
	\label{th:lowerbound}
	The competitive ratio of any deterministic online algorithm for the learning model of Dynamic Balanced Graph Partitioning is at least
	 $(k-2)(\ell-1)/2 - 2$ for any $k\geq 3$ and $\ell \geq 2$.
\end{theorem}

\begin{proof}
	Fix any online algorithm \ALG.
	For a ground set $C$ of nodes that are initially collocated in one cluster,
	let $I(C)$ denote the cluster.
%	If $C$ contains nodes that originate from different clusters
%	then $I(C)$ is undefined.
%	Throughout the construction, we maintain ground sets for each node.
	Initially, all nodes are isolated, i.e., each node is in a singleton ground set.
	First, we choose a cluster arbitrarily and
	 create a~ground set $B$ of $k-1$ nodes in this cluster.
	%We never dismiss any constructed ground set.
	Each cluster hosts exactly $k$ nodes, and in any feasible partition,
	 a~single isolated node must be collocated with $B$.
	At any time,
	we refer to the isolated node currently collocated with $B$ as the \emph{special} node.
	Let $x_0$ denote the first special node.

@@

	We proceed by joining the current special node to a~larger ground set in a similar way,
	as long as a feasible partition exists afterwards.
	Consider the isolated node $x_i$ currently collocated with $B$.
	 We issue a request between $x_i$ and some node $y_i \in C_i$,
	 where $C_i$ is the largest ground set
	 s.t.~$I(C_i) = I(x_i), C_i \neq \{x_0,y_0\}$,
	 only if  a feasible partition exists
	 that collocates all nodes in $\{x_i\} \cup C_i $.
	Any such feasible partition replaces $x_i$ with some isolated node $x_{i+1}$,
	as the new ground set $\{x_i\} \cup C_i $  may not be ever split.
	Otherwise,
	no such feasible partition exists,
	we terminate the process and
	 we let $x^* := x_i, y^*:=y_i$.

	 Once the process terminates,
	 there are at most $\ell$ isolated nodes left 
	 that are not contained in $\{x^*,y^*\}$.
	 Assume for contradiction that this is not the case,
	 i.e., there are at least $\ell+1$ isolated nodes left
	 in addition to $x^*$ and possibly $y^*$.
	 Then there must be two isolated nodes $u$ and $v$,
	 $u,v \not\in \{x^*, y^*\}$,
	  with the same cluster of origin,
	 i.e., $I(\{u\}) = I(\{v\})$.
	 Consider the partition $P$
	  obtained from the initial partition after swapping $x_0$ and $y_0$
	 with $u$ and $v$ (respectively).
	 In this partition,
	 the ground set $\{x_0,y_0\}$ is collocated in the cluster $I(\{u,v\})$.
	 Note that after the first request $\{x_0,y_0\}$,
	   we issue requests only between nodes that have the same cluster of origin and
	   all these nodes are collocated in $P$.
	 Therefore all ground sets constructed so far are collocated in  $P$,
	 i.e., $P$ is a feasible partition,
	 which contradicts our assumption.
	 Hence, at the point of termination,
	 there are at most $\ell$ isolated nodes not in $\{x^*,y^*\}$,
	 or at most $\ell+2$ isolated nodes overall.

\begin{comment}
	We claim that expanding ground sets in this way can continue as long as at least $\ell$ isolated nodes exist.
	Fix the partition just before the creation of a~ground set with $x_i$ for some~$i$.
	We show that if at least $\ell$ isolated nodes exist (including $x_i$), then there exists a~perfect partition of ground sets after joining $x_i$ to a~ground.
	To show this claim, we consider a~partition, where the largest ground set of each cluster originates from it (excluding the ground set $\{x_0, y_0\}$). \mahmoud{this sentence doesn't make sense.}
	In this partition, each cluster contains:  at most one ground set larger than $1$, possibly some isolated nodes, and possibly the ground set $\{x_0, y_0\}$.
	If $I(x_i)$ contains an isolated node, then we simply obtain a~perfect partition by swapping it with $x_i$.
	In the following, we assume that no isolated nodes exist on $I(x_i)$.
	The largest ground set of $I(x_i)$ is at most $k-1$, because $x_i$ is on $S$.
	Therefore $\{x_0, y_0\}$ is present on $I(x_i)$, and we must evict it to make space for $x_i$.
	\mahmoud{why  $\{x_0, y_0\}$ must be present there?}
	To find the cluster for $\{x_0, y_0\}$, observe that we have at most $\ell-1$ clusters with isolated nodes, and we have at least $\ell$ isolated nodes --- hence a~cluster with two isolated nodes exists.
	To find a~feasible partition, we swap $\{x_0, y_0\}$ with these two collocated isolated nodes, and then we swap an isolated node from $I(x_i)$ with~$x_i$.
	\mahmoud{but didn't you assume there is no isolated on $I(x_i)$?}

	To keep \OPT's cost low, we terminate prematurely --- we stop extending ground sets while there are still $\ell+1$ isolated nodes (i.e., neglecting two last ground set expansions).
	\mahmoud{here you are making an exception to the already established process. Instead, adapt the termination condition so that it triggers prematurely, by saying something like "issue the last request only if there will be a feasible partition." }
	Consider \ALG's partition after the last expansion.
	Among $\ell+1$ isolated nodes there must exist a~pair that originates from the same cluster (we have $\ell$ clusters). 
	We denote these two nodes by $x^*$ and $y^*$.
\end{comment}
	The optimal strategy is to perform two node swaps.
	Precisely, \OPT collocates $\set{x_0,y_0}$ 
	by swapping them with $x^*$ and $y^*$.
	In this partition, no ground set is split.
	Due to the termination condition,
	we never issue requests between nodes $x^*, y^*$
	 which are separated in this partition.

	\ALG performs at least one swap each time a~ground set grows,
	i.e., each time it joins the isolated node $x_i$ (collocated with $B$), to  the ground set $C_i$ as specified before.
%	In other words,
%	some ground set grows by one node at each of these steps.
	Consider any ground set $C^*$ at the point of termination.
	This ground set has grown exactly $|C^*|-1$ times until the termination. 
	Let $\mathcal{C}^*$ be the set of all ground sets at the point of termination.
	Thus, $\mathcal{C}^*$ includes ground sets $B$,
	$\{x_0,y_0\}$,
	and up to $\ell+2$ singleton ground sets.
%	(including $x^*$ and possibly $y^*$).	
	Each time we join the special node to the a largest ground set.
	Hence, every  other ground set in $\mathcal{C}^*$ has grown maximally.
	There are at most $\ell-1$ such ground sets in addition to the ones 
	we considered so far,
	one per cluster except the cluster containing $B$. 
	Therefore,  $|\mathcal{C}^*| \leq 1 + 1 + (\ell+2)  + (\ell-1) = 2\ell+3$.
	Note that among all non-singleton ground sets in $\mathcal{C}^*$,
	 only $B$ does not grow during the process.
	 Thus,
    the total number of times that 
	a ground set in $\mathcal{C}^*$  has grown is
	\begin{align*}
	\sum_{C^* \in \mathcal{C}^* } (|C^*|-1) - (k-1) &= 
	\sum_{C^* \in \mathcal{C}^* } |C^*| - 
		\sum_{C^* \in \mathcal{C}^* } 1 - (k-1) \\
	&\geq  k  \ell - (2\ell+3) - (k-1) =  (k-2)(\ell-1) - 4,
	\end{align*}
	which is the least number of swaps performed by \ALG.
	The competitive ratio is then
	$\ALG / \OPT \geq ((k-2)(\ell-1) - 4) / 2$.
	%%
	\begin{comment}
	\ALG performs at least one swap  every time   a~ground set is constructed.
	Each expansion reduces the number of isolated nodes by $1$ or $2$.
	The decrease of two isolated nodes occurs when the largest component on  $I(x_i)$ is still singleton; this occurs at most $\ell$ times.
	We start with $k \cdot \ell - (k-1)$ isolated nodes (a~single $k-1$ ground set is revealed prior to $\{x_0, y_0\}$), and we end with isolated $\ell+1$ nodes.
	Hence, the total number of swaps of \ALG is at least $k\cdot \ell - k - 2 \ell$.
	The competitive ratio is then $\ALG / \OPT \geq (k\cdot \ell - k - 2\ell) / 2$.
	\end{comment}
	%%
\end{proof}

@@

\noindent
\textbf{Perfect Partition Learner algorithm.}
Fix the initial partition
$P_I :=\{ I_1, \dots, I_{\ell}\}$ and \OPT's final partition
$P_F := \{F_1, \dots, F_{\ell}\}$.
The \emph{distance} of a~partition
 $P = \{C_1, \dots, C_{\ell}\}$ from the initial partition,
defined as 
$\Delta(P, P_I) := \sum_{j=1}^{\ell} | C_j \setminus I_j |$,
 is the number of nodes in $P$ that do not reside in their initial cluster.
In other words,
at least $\Delta(P, P_I)$ node migrations are required in order to reach the partition $P$ from $P_I$, and thus
$\OPT \geq \Delta^*:= \Delta(P_F, P_I) $.
We  use $ \Delta(P_F) $ wherever the initial partition is clear from the context \maciek{It always is, we fixed it in the beginning. Maybe just define it as the distance from initial partition, but don't pass it as an argument}.
The scheme of the algorithm can be found in the  appendix (Algorithm \ref{alg:ppl}).

@@

\noindent
\textbf{Approximation for offline rematching.}
\mahmoud{TBD}
%We briefly sketch an approximation for the offline problem,
%that makes use of \RM as a~subroutine.
%In our approximation algorithm,
%we simulate \RM on the given sequence.
%Whenever,
%\RM collocoates a~pair

>>>>

 \usepackage{balance}
 \usepackage{amsthm,amsmath,array,colortbl,graphicx,multirow}
 \usepackage{comment}
 \usepackage{balance}
 \usepackage{tikz}
 \usepackage{amsmath}
 \usetikzlibrary{patterns} %
 \usepackage{algorithm}
 \usepackage[font={footnotesize}]{subcaption}
 \usepackage[font={footnotesize}]{caption}
 \usepackage{breakcites}
 \usepackage{booktabs}
 \usepackage{diagbox}
 \usepackage{xcolor}
 \usepackage{colortbl}
 \usepackage{cleveref}
 \usepackage{enumitem}

@@

\begin{abstract}
	Distributed   applications,  including  batch  processing, streaming, scale-out databases,
	or machine learning, generate a~significant amount of network traffic.
	By collocating frequently communicating nodes (e.g., virtual machines) on the same clusters (e.g., server or rack), we can reduce the network load and  improve application performance. 
	However, the communication pattern of different applications is often unknown a~priori and may change over time, hence it needs to be learned in an online manner.
	%
	This paper revisits the online 
	balanced partitioning problem 
	(introduced by Avin et al.~at DISC 2016)
	that asks for an algorithm that strikes
	an optimal tradeoff between the benefits
	of collocation (i.e., lower network load) 
	and its costs (i.e., migrations). 
	%
	Our first contribution is a~significantly improved deterministic
	lower bound of $\Omega(k\cdot \ell)$ on the
	competitive ratio, where $\ell$ is the number
	of clusters and $k$ is the cluster size,
	even for a~scenario in which the communication
	pattern is static and can be perfectly partitioned;
	we also provide an asymptotically tight upper bound 
	of $O(k\cdot \ell)$ for this scenario.
	For $k=3$, we contribute an asymptotically tight upper bound
	of $\Theta(\ell)$
	for the general model in which the
	communication pattern can change arbitrarily over time.
	We improve the result for $k=2$ by providing a~strictly $6$-competitive upper bound for the general model.
	In contrast to most prior work, our algorithms respect all capacity constraints and do not require resource augmentation.

@@

\begin{theorem}
	\label{th:lowerbound}
	The competitive ratio of any deterministic online algorithm for the learning model of Dynamic Balanced Graph Partitioning is at least
	 $(k-2)(\ell-1)/2 - 3/2$ for any $k\geq 3$ and $\ell \geq 2$.
\end{theorem}

\begin{proof}
	Fix any online algorithm \ALG.
	For a ground set $C$ of nodes that are initially collocated in one cluster,
	let $I(C)$ denote the cluster,
	i.e., the cluster that is the \emph{origin} of $C$.
%	If $C$ contains nodes that originate from different clusters
%	then $I(C)$ is undefined.
%	Throughout the construction, we maintain ground sets for each node.
	Initially, all nodes are isolated, i.e., each node is in a singleton ground set.
	First, we choose a cluster arbitrarily and
	 create a~ground set $B$ of $k-1$ nodes in this cluster.
	%We never dismiss any constructed ground set.
	Each cluster hosts exactly $k$ nodes, and in any feasible partition,
	 a~single isolated node must be collocated with $B$.
	At any time,
	we refer to the isolated node currently collocated with $B$ as the \emph{special} node.
	Let $x_0$ denote the first special node.

@@

	We proceed by joining the current special node to a~larger ground set in a similar way,
	as long as a feasible partition exists afterwards.
	Consider the isolated node $x_i$ currently collocated with $B$.
	 We issue a request between $x_i$ and some node $y_i \in C_i$,
	 where $C_i$ is the largest ground set
	 s.t.~$I(C_i) = I(x_i), C_i \neq \{x_0,y_0\}$,
	 only if  a feasible partition exists
	 that collocates all nodes in $\{x_i\} \cup C_i $.
	Any such feasible partition replaces $x_i$ with some isolated node $x_{i+1}$,
	as the new ground set $\{x_i\} \cup C_i $  may not be ever split.
	Otherwise,
	no such feasible partition exists and
	we terminate the process.

	 Once the process terminates,
	 there are at most $\ell$ isolated nodes left. 
	 Assume for contradiction that this is not the case,
	 i.e., there are at least $\ell+1$ isolated nodes left.
	 Then there must be two isolated nodes $u$ and $v$,
	  with the same cluster of origin,
	 i.e., $I(\{u\}) = I(\{v\})$.
	 Consider the partition $P$
	  obtained from the initial partition after swapping $x_0$ and $y_0$
	 with $u$ and $v$ (respectively).
	 In this partition,
	 the ground set $\{x_0,y_0\}$ is collocated in the cluster $I(\{u,v\})$.
	 Note that after the first request $\{x_0,y_0\}$,
	   we issue requests only between nodes that have the same cluster of origin and
	   all these nodes are collocated in $P$.
	 Therefore all ground sets constructed so far are collocated in  $P$,
	 i.e., $P$ is a feasible partition,
	 which contradicts our assumption.
	 Hence, there are at most $\ell$ isolated nodes
	 at the point of termination.

	The number of isolated nodes decrease after each step.
	Consider the last step, $i^*$, when exactly $\ell+1$ isolated nodes exist before  $x_{i^*}$ joins to another ground set.
	I.e., there are less than $\ell+1$ isolated nodes after the step $i^*$.
	Then, at the end of step $i^*-1$,
	there exist two isolated nodes $x^*$ and $y^*$
	with the same cluster of origin, i.e., $I(\{x^*\})=I(\{y^*\})$.
%
    In what follows,
    we bound costs incurred by \OPT and \ALG over the  sequence of requests
    $\sigma := \{ \{x_0,y_0\}, \dots, \{x_{i^*-1},y_{i^*-1}\}  \}$.

	The optimal strategy is to perform two node swaps.
	Precisely, \OPT collocates $\set{x_0,y_0}$ 
	by swapping them with $x^*$ and $y^*$.
	In this partition, no ground set is split and
%	there is no request to $x^*$ or $y^*$ in $\sigma$.
	therefore \OPT pays only for the two swaps.

	\ALG performs at least one swap each time a~ground set grows,
	i.e., each time it joins the isolated node $x_i$ (collocated with $B$), to  the ground set $C_i$ as specified before.
%	In other words,
%	some ground set grows by one node at each of these steps.
	Consider any ground set $C^*$ after serving $\sigma$.
	This ground set has grown exactly $|C^*|-1$ times until the end of $\sigma$. 
	Let $\mathcal{C}^*$ be the set of all ground sets after serving the last request in $\sigma$.
	Thus, $\mathcal{C}^*$ includes ground sets $B$,
	$\{x_0,y_0\}$,
	and $\ell+1$ singleton ground sets.
%	(including $x^*$ and possibly $y^*$).	
	Each time we join the special node to the largest ground set of same origin.
	Hence, every  other ground set in $\mathcal{C}^*$ has grown maximally.
	There are at most $\ell-1$ such ground sets in addition to the ones 
	we considered so far,
	one per cluster except the cluster containing $B$. 
	Therefore,  $|\mathcal{C}^*| \leq 1 + 1 + (\ell+1)  + (\ell-1) = 2\ell+2$.
	Note that among all non-singleton ground sets in $\mathcal{C}^*$,
	 only $B$ does not grow during the process.
	 Thus,
    the total number of times that 
	a ground set in $\mathcal{C}^*$  has grown is
	\begin{align*}
	\sum_{C^* \in \mathcal{C}^* } (|C^*|-1) - (k-1) &= 
	\sum_{C^* \in \mathcal{C}^* } |C^*| - 
		\sum_{C^* \in \mathcal{C}^* } 1 - (k-1) \\
	&\geq  k  \ell - (2\ell+2) - (k-1) =  (k-2)(\ell-1) - 3,
	\end{align*}
	which is the least number of swaps performed by \ALG.
	The competitive ratio is then
	$\ALG / \OPT \geq ((k-2)(\ell-1) - 3) / 2$.
	%%
	\begin{comment}
	\ALG performs at least one swap  every time   a~ground set is constructed.
	Each expansion reduces the number of isolated nodes by $1$ or $2$.
	The decrease of two isolated nodes occurs when the largest component on  $I(x_i)$ is still singleton; this occurs at most $\ell$ times.
	We start with $k \cdot \ell - (k-1)$ isolated nodes (a~single $k-1$ ground set is revealed prior to $\{x_0, y_0\}$), and we end with isolated $\ell+1$ nodes.
	Hence, the total number of swaps of \ALG is at least $k\cdot \ell - k - 2 \ell$.
	The competitive ratio is then $\ALG / \OPT \geq (k\cdot \ell - k - 2\ell) / 2$.
	\end{comment}
	%%
\end{proof}

@@

\noindent
\textbf{Perfect Partition Learner algorithm.}
Fix the initial partition
$P_I :=\{ I_1, \dots, I_{\ell}\}$ and \OPT's final partition
$P_F := \{F_1, \dots, F_{\ell}\}$.
The \emph{distance} of a~partition
 $P = \{C_1, \dots, C_{\ell}\}$ from the initial partition,
defined as 
$\Delta(P) := \sum_{j=1}^{\ell} | C_j \setminus I_j |$,
 is the number of nodes in $P$ that do not reside in their initial cluster.
In other words,
at least $\Delta(P)$ node migrations are required in order to reach the partition $P$ from $P_I$, and thus
$\OPT \geq \Delta^*:= \Delta(P_F) $.
%We  use $ \Delta(P_F) $ wherever the initial partition is clear from the context
The scheme of the algorithm can be found in the  appendix (Algorithm \ref{alg:ppl}).

@@

\noindent
%\textbf{Approximation for offline rematching.}
%We briefly sketch an approximation for the offline problem,
%that makes use of \RM as a~subroutine.
%In our approximation algorithm,
%we simulate \RM on the given sequence.
%Whenever,
%\RM collocoates a~pair
